---
title: "Analysis for 'Dawnn: single-cell differential abundance with neural networks'"
author: "George T. Hall"
date: "Compiled on `r format(Sys.time(), '%d %B %Y')`"
output:
    html_document:
        toc: false
        code_folding: none
        keep_md: yes
---

<!--
(C) University College London 2023. This software is licenced under the terms
of the GNU GENERAL PUBLIC LICENSE Version 3. See COPYING.txt for the licence
details.
-->

```{r knitr_options, results = 'hide', message = FALSE, warning = FALSE, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.align = "center",
                      fig.height = 14 / 2.54, fig.width = 20 / 2.54, dpi = 300)
```

This notebook generates the figures in the paper _Dawnn: single-cell
differential abundance with neural networks_. It is my intention that all
results shoud be completely reproducible. To this end, all objects and code
used to generate the results files read in this notebook are available
[here](this is the link). If you are unable to reproduce any result, please let
me know at `george.hall@ucl.ac.uk`.

```{r package_imports, message = FALSE, warning = FALSE}
library(tidyverse)
library(ggplot2)
library(scales)
library(plotly)
library(ggpubr)
library(patchwork)
library(rstatix)
f <- glue::glue
set.seed(123)
```

# Benchmarking TPR and FDR

We start by defining some functions to read results files and plot the figures.
For tidyness, these functions are hidden by default. Click on the title to
reveal the code.

<details>
<summary>`read_results` Read results file</summary>
```{r fcn_read_results}
read_results <- function(out_file_name) {
    results <- read.csv(out_file_name, header = FALSE, sep = "\t")
    if (substr(basename(dirname(out_file_name)), 1, 3) == "sim") {
        colnames(results) <- c("method", "family", "tpr_pda", "fpr_pda", "fdr_pda",
                               "tpr_ada", "fpr_ada", "fdr_ada", "upreg_pc1",
                               "upreg_pop", "rep")
    } else {
        colnames(results) <- c("method", "family", "tpr_pda", "fpr_pda", "fdr_pda",
                               "tpr_ada", "fpr_ada", "fdr_ada", "upreg_pc1",
                               "upreg_pop", "rep")
    }

    return(results)
}
```
</details>

<details>
<summary>`parse_results` Parse results dataframe</summary>
```{r fcn_parse_results}
parse_results <- function(results, da_mode) {
    stats_cols <- paste0(c("tpr_", "fpr_", "fdr_"), da_mode)
    to_drop <- paste0(c("tpr_", "fpr_", "fdr_"),
                      ifelse(da_mode == "pda", "ada", "pda"))
    results <- select(results, -one_of(to_drop)) %>%
                tidyr::pivot_longer(cols = stats_cols, names_to = "stat")
    results$hline <- 0
    results[results$stat == paste0("tpr_", da_mode), ]$hline <- 10 # Deliberately outside of graph
    results[results$stat == paste0("fdr_", da_mode), ]$hline <- 0.1
    results$method <- factor(results$method,
                             levels = c("daseq", "cna", "milo", "dawnn_pda", "dawnn_ada"),
                             labels = c("DA-seq", "CNA", "Milo", "Dawnn-G", "Dawnn-L"))
    results$stat <- factor(results$stat,
                           levels = stats_cols,
                           labels = c("TPR", "FPR", "FDR"))
    # Milo paper used natural logarithm here (i.e. log()), but I think log2 is
    # more natural (e.g. more common with DE analysis).
    lfcs <- log2(results$upreg_pc1 / (1 - results$upreg_pc1))
    results$upreg_lfc <- as.factor(round(lfcs, 1))

    results <- subset(results, subset = stat %in% c("TPR", "FDR"))
    #results <- results %>% arrange(upreg_pop, rep)

    return(results)
}
```
</details>

<details>
<summary>`plot_accuracy_results` Plot accuracy results</summary>
```{r fcn_plot_accuracy_results}
da_mode_prettifier <- function(da_mode) {
    return(ifelse(da_mode == "pda", "GDA", "LDA"))
}

plot_accuracy_results <- function(results, da_mode, split_by_rarity = FALSE) {
    # Are we constructing the legend?
    main_plot <- nrow(results) > 1

    plot_list <- list()
    parsed <- parse_results(results, da_mode)
    if (!main_plot) {
        parsed <- head(parsed, 1)
    }

    da_mode_pretty <- da_mode_prettifier(da_mode)
    facet_labels <- labeller(stat = c(TPR = paste0("TPR (", da_mode_pretty, ")"),
                                      FDR = paste0("FDR (", da_mode_pretty, ")")))
    shape <- 21
    size <- ifelse(main_plot, 0.2, 5)
    width <- ifelse(main_plot, 0.2, 0)
    p <- ggplot(parsed, aes(x = upreg_lfc, y = value)) +
        geom_boxplot(aes(color = method), outlier.shape = NA,
                     linewidth = 0.3, alpha = 0.5) +
        geom_hline(aes(yintercept = hline, color = method), alpha = 0.75) +
        geom_jitter(data = subset(parsed, rare_celltype == "Rare"),
                    aes(color = method),
                    fill = "white", shape = shape, stroke = 0.1, size = size,
                    width = width) +
        geom_jitter(data = subset(parsed, rare_celltype == "Common"),
                    aes(fill = method),
                    color = "white", shape = shape, stroke = NA, size = 2 * size,
                    width = width) +
        scale_color_manual(values = METHOD_COLS) +
        scale_fill_manual(values = METHOD_COLS) +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
              axis.title.x = element_text(size = 7),
              axis.text = element_text(size = 6),
              axis.title.y = element_blank(),
              strip.background = element_blank(),
              strip.text = element_text(size = 8),
              legend.position = "none") +
        xlab(bquote(Maximum~log[2] * `-fold` ~ change))

    if (main_plot) {
        p <- p + scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1))
    } else {
        return(p)
    }

    if (split_by_rarity) {
        p <- p + facet_grid(rows = vars(rare_celltype, stat),
                            cols = vars(method),
                            labeller = facet_labels) +
        theme(legend.position = "none")
    } else {
        p <- p + facet_grid(cols = vars(method),
                            rows = vars(stat),
                            labeller = facet_labels)
    }

    plot_list[[da_mode]] <- p

    p <- ggarrange(plotlist = plot_list, nrow = 1,
                   legend = FALSE, common.legend = FALSE)

    return(p)
}
```
</details>

<details>
<summary>`plot_accuracy_results_signif` Plot accuracy results with statistics</summary>
```{r fcn_plot_accuracy_results_signif}
plot_accuracy_results_signif <- function(results, da_mode, tpr_fdr) {
    # If all FDRs are 0 then all differences become 0, which causes problems
    # with the statistical test. We therefore artificially increase the fdrs of
    # 0 for each method by a very small amount to introduce a non-zero
    # difference. This increase is not reflected in the plots.
    for (i in seq_along(unique(results$method))) {
        m <- unique(results$method)[i]
        if (nrow(results[(results$fdr == 0) & (results$method == m), ]) > 0) {
            results[(results$fdr == 0) & (results$method == m), ]$fdr <- (i / 10**9)
        }

        if (nrow(results[(results$tpr == 1) & (results$method == m), ]) > 0) {
            results[(results$tpr == 1) & (results$method == m), ]$tpr <- 1 - (i / 10**9)
        }
    }

    parsed <- parse_results(results, da_mode)
    results_wide <- parsed %>%
                        filter(stat == tpr_fdr) %>%
                        select(c(method, upreg_lfc, upreg_pop, dataset, rep, value)) %>%
                        mutate(pair_id = paste(upreg_lfc, upreg_pop, dataset, rep,
                                               sep = "_")) %>%
                        select(pair_id, method, value) %>%
                        pivot_wider(names_from = method, values_from = value)

    method_pairs <- combn(unique(parsed$method), 2, simplify = FALSE)

    stat_test <- lapply(method_pairs, function(pair) {
      method_1 <- pair[1]
      method_2 <- pair[2]

      # Keep only rows where both methods have a value (complete pairs)
      d <- results_wide %>%
            select(pair_id, all_of(c(method_1, method_2)))

      test <- wilcox.test(x = d[[method_1]], y = d[[method_2]], paired = TRUE,
                          exact = TRUE, conf.int = TRUE)

      tab <- tibble(method_1 = method_1, method_2 = method_2,
                    n_pairs = nrow(d), statistic = test$statistic,
                    estimate = test$estimate, p.value = test$p.value)
      return(tab)
    }) %>%
        do.call(rbind, .)

    stat_test_rev <- stat_test
    stat_test_rev$tmp <- stat_test_rev$method_1
    stat_test_rev$method_1 <- stat_test_rev$method_2
    stat_test_rev$method_2 <- stat_test_rev$tmp
    stat_test_rev$tmp <- NULL
    stat_test_rev$estimate <- -1 * stat_test_rev$estimate

    df <- rbind(stat_test, stat_test_rev) %>%
            select(c("method_1", "method_2", "p.value", "estimate")) %>%
            mutate(p.adj = p.adjust(p.value, method = "bonferroni"))

    if (tpr_fdr == "FDR") {
        # Reverse sign of estimate as we want FDR to be minimized
        df$estimate <- -1 * df$estimate
    }

    df$estimate[df$p.adj > 0.05] <- NA

    da_mode_pretty <- da_mode_prettifier(da_mode)
    plot_title <- paste0("Comparison of ", tpr_fdr, " (", da_mode_pretty, ")")
    method_1_color <- "blue"
    method_2_color <- "red"
    not_signif_color <- "gray"
    p <- ggplot(df, aes(x = method_1, y = method_2, fill = estimate)) +
            geom_tile() +
            scale_fill_gradientn(colors = c(method_2_color, not_signif_color,
                                            method_1_color),
                                 na.value = not_signif_color,
                                 limits = c(-1, 1),
                                 breaks = c(-0.85, 0, 0.85),
                                 labels = c("Method 2", "Neither",
                                            "Method 1")) +
            guides(fill = guide_colorbar(ticks.colour = NA,
                                         label.theme = element_text(hjust = 0.5))) +
            geom_text(data = df[df$p.adj > 0.05, ], aes(x = method_1, y = method_2), label = "ns") +
            labs(title = plot_title, x = "Method 1", y = "Method 2",
                 fill = "Best method") +
            theme_bw() +
            theme(panel.grid = element_blank(), aspect.ratio = 1,
                  axis.title.x = element_text(color = method_1_color),
                  axis.title.y = element_text(color = method_2_color),
                  axis.text.x = element_text(angle = 45, hjust = 1),
                  plot.title = element_text(hjust = 0.5))

    return(p)
}
```
</details>

<details>
<summary>`ggsave_plots` Save ggplots</summary>
```{r ggsave_plots}
ggsave_plots <- function(non_stat_plot, stat_plot, file_name) {
    ggsave(filename = paste0(file_name, ".png", collapse = ""),
           plot = non_stat_plot, dpi = 300, units = "cm", width = 12,
           height = 10)
    ggsave(filename = paste0(file_name, "_stat.png", collapse = ""),
           plot = stat_plot, dpi = 300, units = "cm", width = 24,
           height = 14)
}
```
</details>

<br>

We now plot the accuracy results (with and without tests for statistical
significance) for each benchmarking dataset.

```{r make_results_plots, results = "asis"}
# Generated as in data_for_benchmarking_results/collect_results_all_sim_dat.R
results_prefix <- "benchmarking_rerun_2025/results/"
images_prefix <- "manuscript/images_2025/"
samples <- c("sim_discrete_clusters", "sim_linear_traj", "sim_branch_traj",
             "mouse_embryo", "skin", "organoid", "heart")
# Set to 0 for fully simulated structures as there are no clusters with below
# 8% abundance
rare_pop_cutoff <- list("sim_discrete_clusters" = 0.05, "sim_linear_traj" = 0.05,
                        "sim_branch_traj" = 0.05, "skin" = 0.05,
                        "mouse_embryo" = 0.05, "organoid" = 0.05,
                        "heart" = 0.05)
upreg_pops_proportions <- read.csv("upreg_pops_proportions.csv")

METHOD_COLS <- c("DA-seq" = "#1E88E5", "CNA" = "#004D40", "Milo" = "#FFC107",
                 "Dawnn-G" = "#D81B60", "Dawnn-L" = "#000000")

all_results <- data.frame("method" = character(), "family" = character(),
                          "tpr_pda" = numeric(), "fpr_pda" = numeric(),
                          "fdr_pda" = numeric(), "tpr_ada" = numeric(),
                          "fpr_ada" = numeric(), "fdr_ada" = numeric(),
                          "upreg_pc1" = numeric(), "upreg_pop" = numeric(),
                          "rep" = numeric(), rare_celltype = character(),
                          "dataset" = character())
plot_list_accs <- list("pda" = list(), "ada" = list())
LEGEND_COLS <- sapply(METHOD_COLS, function(x) {"grey"})
for (s in samples) {
    results <- read_results(paste0(results_prefix, s, "/all_methods.csv"))
    pop_prop_df <- subset(upreg_pops_proportions, sample == s)
    rare_celltypes <- subset(pop_prop_df, proportion <= rare_pop_cutoff[[s]])
    rare_celltypes <- rare_celltypes$population
    results$rare_celltype <- ifelse(results$upreg_pop %in% rare_celltypes,
                                    "Rare", "Common")
    results$dataset <- s
    all_results <- rbind(all_results, results)
    for (split_by_rarity in c(TRUE, FALSE)) {
        for (da_mode in c("pda", "ada")) {
            p_acc <- plot_accuracy_results(results, da_mode, split_by_rarity)
            print(p_acc)
            if (!split_by_rarity) {
                plot_list_accs[[da_mode]][[s]] <- p_acc
            }
            suffix_extended <- paste0(ifelse(split_by_rarity, "_split_rarity_", "_"),
                                      da_mode)
            ggsave(filename = paste0(images_prefix, s, suffix_extended, ".pdf",
                                     collapse = ""),
                   plot = p_acc, dpi = 300, units = "cm",
                   width = ifelse(split_by_rarity, 12, 10),
                   height = ifelse(split_by_rarity, 8, 5))
        }
    }

    # Create legend
    pl_leg <- list()
    for (cr in c("Common", "Rare")) {
        for_legend <- subset(results, (rare_celltype == cr)) %>% head(1)
        if (nrow(for_legend) > 0) {
            p <- plot_accuracy_results(for_legend, da_mode, FALSE)
            p$layers[[1]] <- NULL
            p$layers[[1]] <- NULL
            p <- p + theme_void() + theme(legend.position = "none")
            p$layers[[1]]$aes_params$size <- 5
            p$layers[[2]]$aes_params$size <- 7
            p$layers[[1]]$aes_params$stroke <- 2
            p <- p +
                    scale_color_manual(values = LEGEND_COLS) +
                    scale_fill_manual(values = LEGEND_COLS) +
                    annotate("text", label = cr, x = -Inf, y = -Inf, vjust = 2.3,
                             hjust = ifelse(cr == "Common", -0.4, -1.3)) +
                    scale_y_continuous(expand=c(0, 0))
            pl_leg[[cr]] <- p
        }
    }
    p_leg <- ggarrange(plotlist = pl_leg, nrow = 1)
    ggsave(filename = paste0(images_prefix, "tpr_fdr_legend.pdf"), plot = p_leg,
           dpi = 300, units = "cm", width = 6, height = 1.5)

    results <- mutate(results, method = case_when(method == "milo" ~ "Milo",
                              method == "daseq" ~ "DA-seq",
                              method == "dawnn_pda" ~ "Dawnn-G",
                              method == "dawnn_ada" ~ "Dawnn-L",
                              method == "cna" ~ "CNA"))
    default_color <- "black"
    results_idxs <- match(results$upreg_pop, pop_prop_df$population)
    results$upreg_prop <- pop_prop_df[results_idxs, "proportion"]
    results$method <- factor(results$method,
                             levels = c("DA-seq", "CNA", "Milo", "Dawnn-G",
                                        "Dawnn-L"))
    plot_list <- list()
    results <- subset(results, method != "CNA")
    for (statistic in c("tpr", "fdr")) {
        for (da_type in c("ada", "pda")) {
            colname <- paste0(statistic, "_", da_type)
            p <- ggplot(results, aes(x = upreg_prop, y = .data[[colname]], color = method)) +
                    geom_jitter(aes(shape = method), size = 0.75, width = 0.001, stroke = 0.25,
                                alpha = 0.5) +
                    geom_vline(xintercept=0.05, color="#ADADAD",linetype="dashed") +
                    scale_shape_manual(values=c("DA-seq" = 4, "CNA" = 0, "Milo" = 2,
                                                "Dawnn-G" = 16, "Dawnn-L" = 16)) +
                    scale_color_manual(values=c("DA-seq" = default_color, "CNA" = default_color,
                                                "Milo" = default_color,
                                                "Dawnn-G" = "blue",
                                                "Dawnn-L" = "red")) +
                    scale_x_continuous(breaks=scales::pretty_breaks()) +
                    guides(shape = guide_legend(override.aes = list(size = 2))) +
                    xlab("Proportion of cells in upregulated population") +
                    ylab(paste(ifelse(statistic == "tpr", "TPR", "FDR"),
                         ifelse(da_type == "pda","(global DA)", "(local DA)"))) +
                    ylim(0, 1) +
                    theme_bw()
            if (statistic == "fdr") {
                p <- p + geom_hline(yintercept=0.1,color="#ADADAD")
            }
            plot_list[[colname]] <- p
        }
    }
    p <- ggarrange(plotlist = plot_list, common.legend = TRUE, legend = "bottom")
    print(p)
    f_name <- paste0(images_prefix, s, "_no_cna_rare_common.pdf",
                     collapse = "")
    ggsave(filename = f_name, plot = p, dpi = 300, units = "cm", width = 20,
           height = 15)
}

plot_list <- list()
for (tpr_fdr in c("TPR", "FDR")) {
    for (da_mode in c("pda", "ada")) {
        p_name <- paste0(tpr_fdr, "_", da_mode)
        p <- plot_accuracy_results_signif(all_results, da_mode, tpr_fdr)
        print(p)
        plot_list[[p_name]] <- p
    }
}
p <- ggarrange(plotlist = plot_list, common.legend = TRUE, legend = "right")
print(p)
f_name <- paste0(images_prefix, "_tpr_fdr_stats_heatmap.pdf", collapse = "")
ggsave(filename = f_name, plot = p, dpi = 300, units = "cm", width = 20, height = 15)
```

# Recovering liver cirrhosis findings

We follow Dann _et al._ in reproducing published results using Dawnn (see
[Figure 5d](https://www.nature.com/articles/s41587-021-01033-z/figures/5) in
their paper). This section is a very lightly adapted version of [their
code](https://github.com/MarioniLab/milo_analysis_2020/blob/main/notebooks/Fig5_liver_cirrhosis.Rmd#L261).
We demonstrate that DA-seq is also able to recover these published findings.

```{r beeswarm_plotting_functions}
# Label cell types as DA or not, as per dataset's paper
group.by <- "annotation_indepth"
paper_DA <- list(cirrhotic = c("MPs (4)", "MPs (5)", "Endothelia (6)",
                               "Endothelia (7)", "Mes (3)", "Tcells (2)",
                               "Myofibroblasts"),
                 healthy = c("MPs (7)", "Endothelia (1)", "Tcells (1)",
                             "Tcells (3)", "Tcells (1)", "ILCs (1)"))
expDA_df <- bind_rows(
  data.frame(annotation_indepth = paper_DA[["cirrhotic"]],
             pred_DA = "cirrhotic"),
  data.frame(annotation_indepth = paper_DA[["healthy"]],
             pred_DA = "healthy")
  )

create_beeswarm_plot <- function(res, expDA_df) {
    p <- res %>%
      left_join(expDA_df) %>%
      mutate(logFC_color = ifelse(is_da == 1, lfc, NA)) %>%
      arrange(annotation_lineage) %>%
      filter(!is.na(annotation_lineage)) %>%
      ggplot(aes(annotation_indepth, lfc, color = logFC_color)) +
      scale_color_gradient2() +
      guides(color = "none") +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 14)) +
      xlab(group.by) + ylab("Log Fold Change") +
      ggbeeswarm::geom_quasirandom(alpha = 1, size = 0.2) +
      geom_violin(linewidth = NA, scale = "width", bw = 0.2, fill = "#D7D7D7",
                  alpha = 0.5) +
      coord_flip() +
      facet_grid(annotation_lineage~., scales = "free", space = "free") +
      theme_bw(base_size = 22) +
      theme(axis.title.y = element_blank(), axis.text.y = element_blank(),
            axis.ticks.y = element_blank(),
            plot.margin = margin(t = 0, b = 0, l = 0, r = 0, unit = "pt")) +
      ylim(-6.5, 6.5)
}

create_beeswarm_plot_ct_abundance <- function(res, expDA_df, x_col) {
    method_names <- unique(res$method)
    pt_sizes <- rep(1, length(method_names))
    names(pt_sizes) <- method_names
    pt_sizes["milo"] <- ifelse("author_cell_type", 2, 1)

    p <- res %>%
      mutate(logFC_color = ifelse(is_da == 1, lfc, NA)) %>%
      ggplot(aes(y = lfc, x = .data[[x_col]], color = logFC_color)) +
      scale_color_gradient2() +
      scale_size_manual(values = pt_sizes) +
      guides(color = "none") +
      scale_x_discrete(drop = FALSE) +
      scale_y_continuous(breaks = scales::pretty_breaks(n = 7)) +
      ggbeeswarm::geom_quasirandom(size = 0.3, stroke = 0, varwidth = TRUE) +
      geom_violin(linewidth = NA, scale = "width", bw = 0.2, fill = "#D7D7D7",
                  alpha = 0.5) +
      coord_flip() +
      theme_bw(base_size = 22)
  return(p)
}
```

```{r generate_liver_beeswarms, fig.width = 30 / 2.54, fig.height = 40 / 2.54, fig.asp = 2}
# Results generated by "benchmarking_liver_cirrhosis_analysis.R".
liver_cirrhosis_results <- read.csv("liver_analysis_regen/liver_cirrhosis_results_rerun.csv")
ct_nums <- read.csv("liver_analysis_regen/cell_type_abundances.csv")
ct_idxs <- match(liver_cirrhosis_results$annotation_indepth, ct_nums$Var1)
liver_cirrhosis_results$ct_num <- ct_nums[ct_idxs, "Freq"]
liver_cirrhosis_results$ct_prop <- liver_cirrhosis_results$ct_num / sum(ct_nums$Freq)

bio_results <- list()
bio_results[["cirrhosis"]] <- liver_cirrhosis_results

cirrhosis_dfs <- list()
cirrhosis_methods <- c("milo", "daseq", "dawnn_pda", "dawnn_ada")

for (method in cirrhosis_methods) {
    cirrhosis_dfs[[method]] <- liver_cirrhosis_results[liver_cirrhosis_results$method == method, ]
}

# Subset only the cell types also caught by Milo
cell_types_in_milo <- unique(cirrhosis_dfs[["milo"]]$annotation_indepth)
for (method in cirrhosis_methods[cirrhosis_methods != "milo"]) {
    cirrhosis_dfs[[method]] <- subset(cirrhosis_dfs[[method]],
                                      annotation_indepth %in% cell_types_in_milo)
}

pretty_title <- c("milo" = "Milo", "daseq" = "DA-seq", "dawnn_pda" = "Dawnn-G",
                  "dawnn_ada" = "Dawnn-L")
plot_list <- list()
for (method in cirrhosis_methods) {
    plot_list[[method]] <- create_beeswarm_plot(cirrhosis_dfs[[method]], expDA_df) +
                            theme(strip.text.y = element_blank(),
                                  axis.title.x = element_blank()) +
                            ggtitle(pretty_title[method])
}

lhs <- cirrhosis_dfs[["dawnn_pda"]] %>%
  left_join(expDA_df) %>%
  group_by(annotation_indepth) %>%
  summarise(pred_DA = dplyr::first(pred_DA),
            annotation_lineage = dplyr::first(annotation_lineage)) %>%
  mutate(end = ifelse(pred_DA == "healthy", 0, 1),
         start = ifelse(pred_DA == "healthy", 1, 0)) %>%
  filter(!is.na(annotation_lineage)) %>%
  ggplot(aes(annotation_indepth, start, xend = annotation_indepth, yend = end,
             color = pred_DA)) +
  geom_segment(size = 1,
               arrow = arrow(length = unit(0.1, "npc"), type = "closed")) +
  coord_flip() +
  xlab("Cell type annotation") +
  facet_grid(annotation_lineage~., scales = "free", space = "free") +
  scale_color_brewer(palette = "Set1", direction = -1,
                     labels = c("enriched in cirrhotic", "enriched in healthy"),
                     na.translate = FALSE,
                     name = "Ramachandran et al.\nDA predictions") +
  guides(color = guide_legend(ncol = 1)) +
  theme_bw(base_size = 22) +
  ylim(-0.1, 1.1) +
  theme(strip.text.y = element_blank(),
        strip.text.x = element_text(angle = 90),
        plot.margin = unit(c(0, 0, 0, 0), "cm"), panel.grid = element_blank(),
        axis.title.x = element_blank(), axis.text.x = element_blank(),
        axis.ticks.x = element_blank(), legend.position = "bottom")

lhs_legend <- as_ggplot(get_legend(lhs))
lhs <- lhs + theme(legend.position = "none")

p_all_beeswarms <- lhs + plot_list[["milo"]] + plot_list[["daseq"]] +
                    plot_list[["dawnn_pda"]] + plot_list[["dawnn_ada"]] +
                    plot_layout(widths = c(1, 10, 10, 10, 10)) &
                    theme(legend.position = "none",
                          plot.title = element_text(hjust = 0.5))
lfc_legend <- ggplot() +
                annotate("text", size = 8, label = "Log-fold change", x = 0,
                         y = 0) +
                theme_void()
lfc_legend <- ggarrange(ggplot() + theme_void(), lfc_legend, nrow = 1, widths = c(1 , 13))
p_all_beeswarms <- ggarrange(p_all_beeswarms,
                             lfc_legend,
                             lhs_legend, ncol = 1, heights = c(0.9, 0.03, 0.08))
print(p_all_beeswarms)
ggsave(filename = paste0(images_prefix, "cirrhosis_all_methods_beeswarms.png"),
       plot = p_all_beeswarms, dpi = 300, units = "cm", width = 30, height = 40, bg="white")

order_md_col <- function(md, prop_table, ct_col) {
    md$ct_prop <- as.numeric(prop_table[md[, ct_col]])
    ct_ordered <- md %>%
                    select(ct_col, ct_prop) %>%
                    distinct() %>%
                    arrange(ct_prop) %>%
                    na.omit() %>%
                    mutate(combined = paste0(.data[[ct_col]], " [",
                                             format(round(100 * ct_prop, 2),
                                                    nsmall = 2),
                                             "%]"))
    md[, ct_col] <- factor(md[, ct_col], levels = ct_ordered[, ct_col],
                           labels = ct_ordered$combined)
    return(md)
}

# Now make bee swarms with cell type proportion on y axis
plot_list <- list()
prop_table <- prop.table(table(cirrhosis_dfs[["dawnn_ada"]]$annotation_indepth))
for (method in cirrhosis_methods) {
    cirrhosis_dfs[[method]] <- order_md_col(cirrhosis_dfs[[method]],
                                            prop_table, "annotation_indepth")
    plot_list[[method]] <- create_beeswarm_plot_ct_abundance(cirrhosis_dfs[[method]],
                                                             expDA_df,
                                                             "annotation_indepth") +
                            theme(strip.text.y = element_blank(),
                                  axis.title.y = element_blank()) +
                            xlab("Log-fold change") +
                            ggtitle(pretty_title[method])
}

p <- ggarrange(plotlist = plot_list, nrow = 1)
print(p)
ggsave(filename = paste0(images_prefix,
                         "cirrhosis_all_methods_beeswarms_ct_abundance.png"),
       plot = p, dpi = 300, units = "cm", width = 60, height = 30, bg="white")

CELL_TYPES <- c("Terminal airway-enriched secretory cells", "Glial, Schwann cells",
                "Endothelial cells, capillary, common",
                "Endothelial cells, capillary, aerocyte-enriched", "Mast cells",
                "CD8\\+ enriched T, common subtype",
                "Ionocytes")
PATTERN <- paste(CELL_TYPES, collapse = "|")

# Now make beeswarms about biological impact
iono_results <- read.csv("biological_impact/cf_cells_results.csv") %>%
                    na.omit()
# Remove malformed unicode
iono_results$author_cell_type <- gsub("<U+00EF>", "i", fixed = TRUE,
                                      iono_results$author_cell_type)
prop_table <-  prop.table(table(iono_results$author_cell_type))
iono_results <- order_md_col(iono_results, prop_table, "author_cell_type")
bio_results[["iono"]] <- iono_results

iono_dfs <- list()
for (method in cirrhosis_methods) {
    method_results <- iono_results[iono_results$method == method, ]
    iono_dfs[[paste0(method, "_full")]] <- method_results
    idxs <- grepl(PATTERN, method_results$author_cell_type)
    method_results_sub <- method_results[idxs, ]
    method_results_sub$author_cell_type <- factor(method_results_sub$author_cell_type)
    iono_dfs[[paste0(method, "_sub")]] <- method_results_sub
}

max_lfc <- floor(max(unlist(lapply(iono_dfs, function(x) {max(x$lfc)})))+1)
min_lfc <- floor(min(unlist(lapply(iono_dfs, function(x) {min(x$lfc)}))))

for (full_or_sub in c("_full", "_sub")) {
    plot_list <- list()
    x_lab <- paste0("Cell type", ifelse(full_or_sub == "_full", " ", "\n"),
                    "[percentage of dataset]")
    for (method in cirrhosis_methods) {
        to_plot <- iono_dfs[[paste0(method, full_or_sub)]]
        p <- create_beeswarm_plot_ct_abundance(to_plot, expDA_df,
                                               "author_cell_type") +
                theme(axis.text.y = element_text(size = 10)) +
                ylim(min_lfc, max_lfc) +
                xlab(x_lab) +
                ylab("Log-fold change") +
                ggtitle(pretty_title[method])
        p$layers <- rev(p$layers)
        plot_list[[method]] <- p
    }

    p <- ggarrange(plotlist = plot_list, nrow = 2, ncol = 2)
    print(p)
    p_name <- paste0(images_prefix,
                     f("ionocytes_all_methods_beeswarms_ct_abundance{full_or_sub}.pdf"))
    ggsave(filename = p_name, plot = p, dpi = 300, units = "cm", width = 50,
           height = ifelse(full_or_sub == "_full", 60, 20), bg = "white")
}
```

```{r da_calls_vs_cell_type_proportion}
# Drop NAs from Milo's results (not sure why they are being generated in the first place?). Plot proportion of cells (or in Milo's case neighbourhoods) called as DA in for each cell type.
for (dataset in c("cirrhosis", "iono")) {
    df <- bio_results[[dataset]] %>%
            na.omit()

    if (dataset == "cirrhosis") {
        df <- df %>%
                group_by(method, annotation_indepth, ct_prop)
    } else if (dataset == "iono") {
        df <- df %>%
                group_by(method, author_cell_type, ct_prop)
    } else {
        stop("Unrecognized dataset: ", dataset)
    }

    df <- df %>%
            mutate(method = factor(method,
                                   levels = c("daseq", "cna", "milo",
                                              "dawnn_pda", "dawnn_ada"),
                                   labels = c("DA-seq", "CNA", "Milo",
                                              "Dawnn-G", "Dawnn-L"))) %>%
            summarize(pct_da = sum(is_da) / n(), pct_pos_da = sum(is_da & (lfc > 0)) / n())

    for (da_type in c("pct_da", "pct_pos_da")) {
        p <- ggplot(df, aes(x = ct_prop, y = .data[[da_type]], color = method)) +
                geom_line(alpha = 0.3) +
                geom_point(size = 0.5) +
                xlab("Cell type proportion") +
                ylab(paste0("Proportion of cells called as\n",
                            ifelse(da_type == "pct_pos_da", "positively ", ""),
                            "differentially abundant")) +
                labs(color = "Method") +
                theme_bw()
        print(p)
        ggsave(filename = paste0(images_prefix, dataset, "_", da_type,
                                 "_vs_ct_abundance.png"),
               plot = p, dpi = 300, units = "cm", width = 20, height = 7, bg="white")
    }
}

# Now plot the amount of DA missed by each global method:
ct_col <- ifelse(dataset == "cirrhosis", "annotation_indepth", "author_cell_type")
p <- pivot_wider(df, names_from = method, values_from = pct_da) %>%
        mutate(`Dawnn-L` - across(c(`DA-seq`, `Dawnn-G`, Milo)), .keep = "unused") %>%
     pivot_longer(-c(.data[[ct_col]], ct_prop), names_to = "method",
                  values_to = "pct_missed") %>%
     ggplot(aes(x = ct_prop, y = pct_missed, color = method)) +
         geom_line(alpha = 0.3) +
         geom_point(size = 0.5) +
         ylim(c(-1, 1)) +
         xlab("Cell type proportion") +
         ylab("Proportion more cells\ncalled as DA by Dawnn-L") +
         labs(color = "Method") +
         theme_bw()
print(p)
ggsave(filename = paste0(images_prefix,
                         "cirrhosis_pct_da_vs_ct_abundance_vs_dawnn_L.png"),
       plot = p, dpi = 300, units = "cm", width = 20, height = 7, bg="white")
```


```{r proportion_called_da}
plot_proportion_called_da <- function(results, plot_filename, cell_type_name) {
    tab <- table(results$is_da,
           results[[cell_type_name]],
           results$method)["TRUE", , ]
    ct_freqs_long <- tab %>%
              data.frame() %>%
              rename(all_of(c("cell_type" = "Var1", "method" = "Var2",
                     "freq" = "Freq"))) %>%
              filter(method != "milo") %>%
              mutate(method = case_when(method == "milo" ~ "Milo",
                            method == "daseq" ~ "DA-seq",
                            method == "dawnn_pda" ~ "Dawnn-G",
                            method == "dawnn_ada" ~ "Dawnn-L"))

    for (plot_type in c("", "_log")) {
        p <- ggplot(ct_freqs_long, aes(x = cell_type, y = freq, fill = method)) +
             geom_bar(position = "dodge", stat = "identity") +
             scale_y_continuous(trans = ifelse(plot_type == "_log", "log10",
                                               "identity")) +
             scale_fill_manual(values = METHOD_COLS) +
             xlab("Cell type") +
             ylab("Number of DA cells") +
             theme_bw() +
             theme(plot.title = element_text(hjust = 0.5),
                   legend.title = element_blank(),
                   axis.text.x = element_text(angle = ifelse(cell_type_name == "annotation_lineage",45,90), hjust = 1))
        print(p)
        ggsave(filename = paste0(plot_filename, plot_type, ".pdf"), plot = p, dpi = 300,
               units = "cm",
               width = ifelse(cell_type_name == "annotation_lineage", 15, 30),
               height = ifelse(cell_type_name == "annotation_lineage", 8, 12))
    }

    ct_freqs_wide <- ct_freqs_long %>%
                        pivot_wider(names_from = method, values_from = freq) %>%
                        mutate(across(c("Dawnn-L", "Dawnn-G"),
                                  ~ paste0(comma(.),
                                           " (", round(100*(./`DA-seq`))-100, "\\%)")
                        )) %>%
                        mutate(`DA-seq` = comma(`DA-seq`))
    old_width <- options("width")$width
    options(width = 250)
    write.table(ct_freqs_wide, sep = " & ", quote = FALSE, row.names = FALSE,
                col.names = FALSE, eol = " \\\\ \n")
    options(width = old_width)
}

plot_proportion_called_da(bio_results[["cirrhosis"]],
                          paste0(images_prefix, "liver_discovery_rates"),
                          "annotation_lineage")
levels(bio_results[["iono"]]$author_cell_type) <- gsub("%", "\\%",
                                                       levels(bio_results[["iono"]]$author_cell_type))
plot_proportion_called_da(bio_results[["iono"]],
                          paste0(images_prefix, "copd_discovery_rates"),
                          "author_cell_type")
```

# Runtime analysis

We compare the runtimes of Dawnn, Milo, and DA-seq.

```{r runtime_results, fig.width = 20 / 2.54, fig.height = 10 / 2.54}
runtime_results <- read.csv("benchmarking_rerun_2025/results/fully_sim/runtime_results_partial_daseq.csv",
                            header = FALSE, sep = "\t")
colnames(runtime_results) <- c("Method", "n", "t", "rep")
runtime_results$Method[runtime_results$Method == "dawnn_pda"] <- "Dawnn-G"
runtime_results$Method[runtime_results$Method == "dawnn_ada"] <- "Dawnn-L"
runtime_results$Method[runtime_results$Method == "milo"] <- "Milo"
runtime_results$Method[runtime_results$Method == "daseq"] <- "DA-seq"
runtime_results$Method[runtime_results$Method == "cna"] <- "CNA"
runtime_results$Method <- factor(runtime_results$Method,
                                 levels = c("DA-seq", "CNA", "Milo", "Dawnn-G",
                                            "Dawnn-L"))
runtime_results$t <- runtime_results$t / 60
means <- runtime_results %>%
            group_by(Method, n) %>%
            summarize(mean = mean(t))

runtime_plot <- ggplot(runtime_results, aes(x = n/1000, y = t, color = Method)) +
                geom_jitter(size = 1.5, alpha = 0.5, stroke = NA, width = 0.5) +
                geom_line(data = means, aes(x = n / 1000, y = mean, color = Method)) +
                xlab("Number of cells (1000s)") +
                ylab("Time (minutes)") +
                scale_y_continuous(breaks = seq(0, 2 * max(runtime_results$t),
                                                by = 30)) +
                scale_color_manual(values = METHOD_COLS) +
                theme_classic() + theme(legend.title = element_blank())
print(runtime_plot)
ggsave(filename = paste0(images_prefix, "runtime_plot.pdf"), plot = runtime_plot,
       dpi = 300, units = "cm", width = 15, height = 10)
```

# Effect of cluster size

```{r effect_cluster_size}
get_total_l1 <- function(y) {
    y + (2 * 0.5 * num_cells_cluster_1_2)
}
get_total_cells <- function(y) {
    y + (2 * num_cells_cluster_1_2)
}

num_cells_cluster_1_2 <- 100
df <- data.frame(num_cells_clust_3 = seq(10, 1000, 2)) %>%
        cbind(num_L1_cells_clust_3 = round(0.6 * .$num_cells_clust_3)) %>%
        cbind(prop_L1_cells_overall = sapply(.$num_L1_cells_clust_3,
                                             get_total_l1) /
                                      sapply(.$num_cells_clust_3,
                                             get_total_cells)) %>%
        cbind(prop_L1_cells_clust_3 = .$num_L1_cells_clust_3 /
                                      .$num_cells_clust_3) %>%
        cbind(LDA = .$prop_L1_cells_clust_3 - 0.5) %>%
        cbind(GDA = .$prop_L1_cells_clust_3 - .$prop_L1_cells_overall) %>%
        pivot_longer(cols = c("GDA", "LDA"),
                     names_to = "score_type", values_to = "score")

p <- ggplot(df, aes(x = num_cells_clust_3, y = score, color = score_type)) +
        geom_line() +
        scale_x_continuous(expand=c(0,0)) +
        scale_y_continuous(expand=c(0,0)) +
        scale_color_manual(values=c("GDA"="#D81B60", "LDA"="#1E88E5")) +
        ylim(0, 0.15) +
        xlab("Number of cells in Cluster 3") +
        ylab("Score") +
        geom_hline(yintercept=0.05, ) +
        theme_bw() +
        theme(legend.title = element_blank(), panel.grid = element_blank())
print(p)

ggsave(paste0(images_prefix, "effect_class_imbalance_plot.pdf"), p, height = 2)
```

# Session info

```{r session_info}
sessionInfo()
```
